---
title: "Global Neonatal Deaths"
author: "Varaidzo Moyo"
format: 
  dashboard
theme: solar
execute:
  echo: false
  warning: false
  message: false
---

# Neonatal Deaths 2021

## Row 1 {height=25%}

The first 28 days of life – the neonatal period – is the most vulnerable time for a child’s survival. Children face the highest risk of dying in their first month of life at an average global rate of 17 deaths per 1,000 live births in 2022, down by 53 per cent from 37 deaths per 1,000 live births in 1990. Globally, 2.3 million children died in the first month of life in 2022 – approximately 6,300 neonatal deaths every day.
source::<https://data.unicef.org/topic/child-survival/neonatal-mortality/>

## Row 2 {height=60%}

```{r}
# Loading packages 
if(!require(pacman)) install.packages("pacman")
pacman::p_load(rlang, tidyverse, knitr,maps, here, reactable,rnaturalearth,ggspatial,mdsr,terra,spData,readr, gt, flextable,dplyr,tidyr,ggplot2,plotly,htmltools,orca)

```

```{r}
#Loading data
neonatal_deaths<- read.csv(here("data/Number_of_neonatal_deaths.csv"))

govt_health_spending<-read.csv(here("data/CHE_gross_domestic_product(GDP).csv"))
```

```{r}
# Selecting only desired columns for Government Health Spending
gvt_health_spending_data <- govt_health_spending%>%
  select(Indicator, Location,SpatialDimValueCode, Period, Value)

#Converting data from long to wide format
wide_data <- gvt_health_spending_data %>%
  arrange(Location) %>%
  pivot_wider(
    id_cols = Location,
    names_from = Period,
    values_from = Value
  )

# Selecting only desired columns for Neonatal deaths
neonatal_data <- neonatal_deaths%>%
  select(Indicator, Location,SpatialDimValueCode, Period, FactValueNumeric)

#Converting data from long to wide format
wide_data_neonatal <- neonatal_data %>%
  arrange(Location) %>%
  pivot_wider(
    id_cols = Location,
    names_from = Period,
    values_from = FactValueNumeric
  )
```

```{r}
#joining data (the Neonatal and Government Expenditure datasets)
gvt_neonatal_joined<-
inner_join(neonatal_data, gvt_health_spending_data, by = c("Location", "Period"))

# Renaming columns in gvt_neonatal_joined
colnames(gvt_neonatal_joined)[colnames(gvt_neonatal_joined) == "FactValueNumeric"] <- "Deaths"

colnames(gvt_neonatal_joined)[colnames(gvt_neonatal_joined) == "Value"] <- "GDP%"

# Filtering data to select 2021
neonatal_deaths_2021 <- gvt_neonatal_joined[gvt_neonatal_joined$Period == 2021, ]
```

```{r}
# Loading world map data
countries <- rnaturalearth::ne_countries()

# Renaming 'Location' column to 'name_long' in gvt_neonatal_joined
colnames(neonatal_deaths_2021)[colnames(neonatal_deaths_2021) == "Location"] <- "name_long"

# Renaming 'Value' column to 'Deaths' in neonatal_deaths_2021
colnames(neonatal_deaths_2021)[colnames(neonatal_deaths_2021) == "Value"] <- "Deaths"

# Merging neonatal deaths data for 2021 with world map data
merged_data <- merge(countries, neonatal_deaths_2021,by = "name_long", all.x = TRUE)


# Creating horopleth map of neonatal deaths
Deaths_map <- ggplot(data = merged_data, aes(fill = Deaths)) +
  geom_sf() +  # Plotting the spatial polygons
  scale_fill_viridis_c(name = "Number of Deaths", labels = scales::comma, na.value = "grey90",
                       option = "plasma", direction = -1) +  # Color scale with comma-separated labels
  labs(fill = "Neonatal Deaths") +  # Legend title
  theme_minimal() +  # Minimal theme
  theme(
    panel.background = element_blank(),  # Removing panel background
    panel.grid.major = element_blank(),   # Removing major grid lines
    panel.grid.minor = element_blank(),   # Removing minor grid lines
    plot.title = element_text(size = 20), # Changing title size
  )


# Save the plot as an image file (e.g., PNG or PDF)
ggsave("C:/Users/Varaidzom/Documents/R training/final_project/Images/global_neonatal_deaths.png", plot = Deaths_map, width = 10, height = 8, dpi = 300)

# Convert ggplot to plotly object
neonatal_deaths <- ggplotly(Deaths_map)

# Display the interactive plot
neonatal_deaths

```
## Row 3{height=15%}

### {width=30%}
```{r}
# Check for NA values in Deaths column
has_na <- sum(is.na(merged_data$Deaths))

if (has_na > 0) {
  # Remove NA values in Deaths column
  merged_data <- merged_data[!is.na(merged_data$Deaths), ]
}

# Calculate highest number of deaths
max_deaths <- max(merged_data$Deaths)
country_max_deaths <- merged_data$name_long[which.max(merged_data$Deaths)]

# Printing results
cat("Country with the highest deaths:", country_max_deaths, "\n")
cat(max_deaths, "\n\n")

```

### {width=30%}
```{r}
# Calculate lowest number of deaths
min_deaths <- min(merged_data$Deaths)
country_min_deaths <- merged_data$name_long[which.min(merged_data$Deaths)]

# Printing results
cat("Country with the lowest deaths:", country_min_deaths, "\n")
cat(min_deaths, "\n\n")
```

### {width=30%}

```{r}
# Calculate average number of deaths
average_deaths <- mean(merged_data$Deaths)

# Printing results
cat("Average number of deaths Globally:", average_deaths, "\n")
```


# 10 year Health expenditure

## Row 1 {height=40%}
General government expenditure on health as a percentage of total government expenditure is the level of general government expenditure on health (GGHE) expressed as a percentage of total government expenditure. It shows the weight of public spending on health within the total value of public sector operations. This indicator includes not just the resources channelled through government budgets, but also the expenditures channelled through government entities for health by parastatals, extrabudgetary entities and, notably, compulsory health insurance.

Health expenditure includes all expenditures for the provision of health services, family planning activities, nutrition activities and emergency aid designated for health, but it excludes the provision of drinking water and sanitation

source::https://www.who.int/data/nutrition/nlis/info/health-expenditure


## Row 2 {height=60%}

### {width=50%}

```{r}
# Listing countries to display
countries_to_display <- c("Zimbabwe", "Afghanistan", "United States of America", "Haiti", "Turkiye","Japan","South Africa","Switzerland","China")

# Filtering neonatal_deaths_2021 to include only the selected countries and years from 2010 to 2021
filtered_data <- gvt_neonatal_joined %>%
  filter(Location %in% countries_to_display,
         Period >= 2010 & Period <= 2021)

# Creating ggplot object
Neonatal_Deaths <- ggplot(filtered_data, aes(x = Period, y = Deaths, color = Location, group = Location)) +
  geom_line() +
  labs(x = "Year",
       y = "Number of Neonatal Deaths",
       title = "10 Year Neonatal Deaths Trend") +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 12),  # Adjusting title font size
    axis.title.x = element_text(size = 10),  # Adjusting x-axis label font size
    axis.title.y = element_text(size = 10),  # Adjusting y-axis label font size
    axis.text.x = element_text(size = 8),    # Adjusting x-axis tick label font size
    axis.text.y = element_text(size = 8)     # Adjusting y-axis tick label font size
  )

# Save the plot as an image file (e.g., PNG or PDF)
ggsave("C:/Users/Varaidzom/Documents/R training/final_project/Images/neonatal_deaths.png", plot = Neonatal_Deaths, width = 10, height = 8, dpi = 300)

# Converting ggplot to plotly for interactivity
neonatal_deaths <- ggplotly(Neonatal_Deaths)

# Displaying the interactive plot
neonatal_deaths
```

### {width=50%}

```{r}
# Loading world map data
countries <- rnaturalearth::ne_countries()

# Renaming 'Location' column to 'name_long' in gvt_neonatal_joined
colnames(neonatal_deaths_2021)[colnames(neonatal_deaths_2021) == "Location"] <- "name_long"

# Renaming 'Value' column to 'Number_of_deaths' in neonatal_deaths_2021
colnames(neonatal_deaths_2021)[colnames(neonatal_deaths_2021) == "Value"] <- "Number_of_deaths"


# Merging neonatal deaths data for 2021 with world map data
merged_data <- merge(countries, neonatal_deaths_2021, by = "name_long", all.x = TRUE)

# Creating choropleth map
GDP_map <- ggplot(data = merged_data, aes(fill = `GDP%`)) +
  geom_sf() +  # Plotting the spatial polygons
  scale_fill_viridis_c(name = "Health Expenditure", na.value = "grey90",
                       option = "viridis", direction = -1) +  # Color scale
  labs(fill = "Health Expenditure (% of GDP)") +  # Legend title
  theme_minimal()  # Minimal theme

# Displaying the map
GDP_map


# Save choropleth map as PNG
ggsave("C:/Users/Varaidzom/Documents/R training/final_project/Images/GDP_map.png", GDP_map, width = 10, height = 8, dpi = 300)

```

# Neonatal deaths vs GDP Expenditure  

## Row 1 {height=70%}

```{r title= "Neonatal deaths vs GDP Expenditure 2021"}
# Renaming 'name_long' column to 'Country' in neonatal_deaths_2021
colnames(neonatal_deaths_2021)[colnames(neonatal_deaths_2021) == "name_long"] <- "Country"

# Creating ggplot object for bubble plot with aggregated data
bubble_plot <- ggplot(neonatal_deaths_2021, aes(x = `GDP%`, y = Deaths,size= Deaths, fill = Country)) +
  geom_point(alpha = 0.8, shape = 21) +  # Filled circles with transparency and white outline
  scale_size(range = c(1, 20), labels = scales::comma) +  # Adjusting bubble size range and format labels as comma-separated integers
  labs(
    x = "Government Health Expenditure (% of GDP)",
    y = "Neonatal Deaths",
    size = "Neonatal Deaths"
  ) +
  theme_minimal(base_size = 10) +  # Minimal theme with base font size
  theme(
    plot.title = element_text(size = 10),  # Adjusting title font size
    plot.background = element_rect(fill = "white"),  # White background
    axis.line = element_line(color = "black")  # Color of axis lines
  ) +
  scale_y_continuous(labels = scales::comma) +  # Formatting y-axis labels as comma-separated integers
  guides(size = FALSE, fill = FALSE)  # Removing size and fill legends

# Save bubble as PNG
ggsave("C:/Users/Varaidzom/Documents/R training/final_project/Images/Neonatal_GDP_plot.png", bubble_plot, width = 10, height = 8, dpi = 300)

# Converting ggplot to plotly for interactivity
interactive_plot <- ggplotly(bubble_plot, tooltip = c("Country", "GDP%", "Deaths"))

# Displaying the interactive plot
interactive_plot
```

## Row 2 {height=30%}
According to WHO, government health expenditure as a percentage of GDP plays a crucial role in improving health outcomes, including reducing neonatal deaths. Effective allocation and utilization of health resources are key to strengthening health systems and achieving better maternal and child health outcomes globally.

Adequate and well-targeted government health expenditure contributes to improving maternal and child health outcomes, including reducing neonatal deaths.
Investments in maternal and child health services, such as prenatal care, skilled birth attendance, postnatal care, immunizations, and newborn care, are crucial for preventing neonatal deaths.




# Download Data

The data used in this dashboard is shown below and can be downloaded as a CSV.

```{r}
library(htmltools)

htmltools::browsable(
  tagList(
    reactable(gvt_neonatal_joined,
      elementId = "neonatal-table",
      searchable = TRUE,
      filterable = TRUE),
    
    tags$button("Download as CSV", onclick = "Reactable.downloadDataCSV('neonatal-table')")
  )
)
```

